# Accuracy, Error, and Precision

So far, we have discussed many different data-related limitations to using geographic information for spatial analysis, such as resolution, abstraction (and other limits of the layer model), data bias, and the impacts of "cooked" data.

We also need to examine how the processes of measurement and representation introduce accuracy issues, errors, and varying levels of precision into our datasets.

The layer approach necessarily relies on abstraction: the process of simplifying complex real-world features into layer representations that GIS software can manage and analyze. Trying to represent every feature with complete accuracy is impossible from a data storage, collection, and analysis perspective

Abstraction always involves decisions, and those decisions will impact the outcome of any GIS analysis using the data. Abstraction is guided by the original purpose of the dataset. The purpose allows data creators to determine what aspects of the real-life feature are important (and therefore should be collected and represented with high accuracy) and which aspects are less important (and can be generalized or omitted). A mismatch between your purpose for using the data and the abstraction decisions made in the data creation process can be a serious issue for the reliability of your results.

Consider the following example of different uses for the same dataset- hospitals in North Carolina:

-   An engineer assessing flood risk at hospitals in North Carolina
-   A public health analyst investigating disease outbreaks across hospitals in North Carolina

What is considered important to these two users differ greatly. The engineer may require detailed building footprints to evaluate structural exposure. The public health analyst may want detailed information on the number of beds and the infectious disease protocols, but just need a rough location for the hospital. If these users were creating the datasets themselves, they would likely look very different. However, most of the time, users aren’t collecting or creating the data themselves, they’re working with data that already exists. As a result, it’s crucial for users to critically evaluate the dataset’s underlying abstraction decisions regarding:

-   Selection - What features are included or excluded?
-   Classification - How are features grouped?
-   Simplification - How much detail of the original feature is retained?
-   Resolution- What is the smallest feature the data can meaningfully represent? 

Furthermore, all spatial data comes with some degree of uncertainty and error

-   Errors include positional inaccuracies, attribute inaccuracies, or conceptual inaccuracies
-   Uncertainty refers more broadly to the limits of our knowledge about the data. This might include incomplete data, ambiguous classifications, unknown data sources, or the inability to verify how a measurement was taken. 

What ultimately matters is not whether the data is perfect, but whether the level and type of abstraction, error, and uncertainty are acceptable enough for the analysis you are doing.

MAUP

## Case Studies

### Floodplain Management in Chapel Hill

The Town of Chapel Hill is interested in analyzing flood risk to buildings across town. They produce two different interactive maps to present to the town council. Look at the maps below and answer the following questions:

-   These maps both show buildings across town as well as the floodplain (flood risk areas). How do the maps differ?
-   How would these differences impact the outcome of an analysis of building flood risk?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(leaflet)
library(sf)
library(htmltools)

# Load spatial data
fp_detailed <- st_read("data/floodplain.geojson", quiet = TRUE)
fp_simple <- st_read("data/simplified_floodplain.geojson", quiet = TRUE)
bldg_polys <- st_read("data/building_footprints.geojson", quiet = TRUE)
bldg_pts <- st_read("data/building_centroids.geojson", quiet = TRUE)

# Map 1a: Detailed
map1_right <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = fp_detailed, color = "blue", weight = 1.5, fillOpacity = 0.3) |>
  addPolygons(data = bldg_polys, color = "black", weight = 0.5) |>
  setView(lng = -79.05, lat = 35.91, zoom = 14)

# Map 1b: Simplified
map1_left <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = fp_simple, color = "blue", weight = 1.5, fillOpacity = 0.3) |>
  addCircleMarkers(data = bldg_pts, radius = 1, color = "black", stroke = FALSE) |>
  setView(lng = -79.05, lat = 35.91, zoom = 14)

library(leafsync)
browsable(sync(map1_left, map1_right, sync.cursor = TRUE))

```

### Building Age

The North Carolina Department of Environmental Quality is considering doing outreach for the Energy Saver NC program. They want to target areas in Orange County that may have aging infrastructure and have data on the average year that homes were built in different census enumeration units. Explore the maps below and ask yourself the following questions:

-   These maps both show maximum heat. How do the maps differ?
-   How would these differences impact the outcome of an analysis of average home age?
-   What key geographic concept is this case study displaying?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Read and prep data
bg <- st_read("data/bg_building_age.geojson", quiet = TRUE)
tract <- st_read("data/tract_building_age.geojson", quiet = TRUE)

bg$SE_A10057_ <- as.numeric(bg$SE_A10057_)
tract$SE_A10057_ <- as.numeric(tract$SE_A10057_)

# Shared color scale
pal <- colorNumeric("YlOrRd", domain = c(1950, 2015), na.color = "#cccccc")

# Map 2a: Block Group
map2_left <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    data = bg,
    fillColor = ~pal(SE_A10057_),
    fillOpacity = 0.6,
    color = "#444", weight = 0.5,
    label = ~paste("BG Year:", SE_A10057_)
  ) |>
  addLegend(pal = pal, values = bg$SE_A10057_, title = "Avg. Year Built") |>
  setView(lng = -79.05, lat = 35.91, zoom = 12)

# Map 2b: Tract
map2_right <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    data = tract,
    fillColor = ~pal(SE_A10057_),
    fillOpacity = 0.6,
    color = "#444", weight = 0.5,
    label = ~paste("Tract Year:", SE_A10057_)
  ) |>
  addLegend(pal = pal, values = tract$SE_A10057_, title = "Avg. Year Built") |>
  setView(lng = -79.05, lat = 35.91, zoom = 12)

# Display synced maps
browsable(sync(map2_left, map2_right, sync.cursor = TRUE))

```

### July Heat Wave

In July 2025 there was a heat wave across the southeast US. A nonprofit group that is focused on planting trees in areas impacted by the urban heat island affect is given two raster datasets that display maximum land surface temperature (LST) on the hottest day of the heatwave. The nonprofit plans to use this data to identify specific streets across Orange County where more trees should be planted. Explore the maps below and ask yourself the following questions

-   These maps both show maximum heat. How do the maps differ?
-   How would these differences impact the outcome of an analysis identifying tree planting priority areas?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(terra)

library(leaflet)
library(terra)
library(leafsync)
library(htmltools)

# Load rasters
high_res <- rast("data/high_res_temp.tif")
low_res <- rast("data/low_res_temp.tif")

# Force min/max calculation
high_range <- range(values(high_res), na.rm = TRUE)
low_range <- range(values(low_res), na.rm = TRUE)

# Use combined range for color palette
combined_range <- range(c(high_range, low_range))

pal <- colorNumeric(
  palette = "inferno",  # requires viridisLite
  domain = combined_range,
  na.color = "transparent"
)

# Map 3a: High-res
map3_left <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addRasterImage(high_res, colors = pal, opacity = 0.6, project = TRUE) |>
  addLegend(pal = pal, values = high_range, title = "Temperature") |>
  setView(lng = -79.05, lat = 35.91, zoom = 9)

# Map 3b: Low-res
map3_right <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addRasterImage(low_res, colors = pal, opacity = 0.6, project = TRUE) |>
  addLegend(pal = pal, values = low_range, title = "Temperature") |>
  setView(lng = -79.05, lat = 35.91, zoom = 9)

browsable(sync(map3_left, map3_right, sync.cursor = TRUE))

```

## Accuracy and Precision

Accuracy can be defined as the degree to which our data matches "true" values. Precision refers to the level of measurement and exactness of our data. A dataset can be:

-   Accurate but not precise- the measurements are close to the "true" value, but recorded at coarse increments

-   Precise but not accurate- the measurements are highly detailed but consistently off from the true value.

For example, let's consider temperature data collected at the RDU ASOS station. In this dataset, accuracy would be defined by how close the value measured at the station is to the actual temperature outside. Precision would be, essentially, how many decimal places there are. For instance, is the station measuring to the degree, the half degree, the tenth of a degree?

## Data Quality and Error

Accuracy and precision then inform data quality– which is the relative accuracy and precision of a particular dataset. Error refers to the combined effects of inaccuracies and imprecision

Error in a geographic dataset can take several different forms:

-   **Positional error**- inaccuracies or imprecision in location. Can be caused by:
    -   Measurement limitations- GPS inaccuracy, sensor resolution limits, survey instrument errors
    -   Map scale- smaller scales generalize more, reducing positional detail
    -   Projection and transformation- coordinate system changed introduce displacement
    -   Topological errors
-   **Attribute error**- inaccuracies or imprecision in the non-spatial data associated with features. Can be caused by:
    -   Data entry or transcription mistakes
    -   Outdated data- attributes no longer represent current conditions
    -   Sensor/device errors- faulty readings from measurement device
    -   Sensor/device limitations- precision or range of measurement is inherently restricted
-   **Conceptual error**- Mismatches between the data's abstraction and the actual characteristics of what it represents. Can be caused by:
    -   Inappropriate classification- categories too broad or narrow
    -   Overgeneralization- simplifying complex features beyond what's appropriate for analysis
    -   Temporal mismatch- datasets from different dates are combined
    -   Modeling assumptions

### When do issues become errors:

Some problems are always errors– for example, a malfunctioning GPS unit or broken temperature sensor will always produce inaccurate data regardless of how the data is used.

Other potential issues only become errors in certain contexts. For instance:

-   Imprecise data is not inherently "wrong", but if your analysis requires high precision, that imprecision becomes an error
-   Low resolution is not automatically an error, it just limits the detail of the data. However, if your research question demands final spatial detail, low resolution would be a source of error.

These are examples of context-dependent error, which are situations where the data's characteristics (precision, resolution, scale) are mismatched with the requirements of the analysis and, therefore, become error.

### Error vs. Uncertainty

Uncertainty describes the estimated amount of error that might be present in a dataset. It allows users to judge whether the data is suitable for a specific purpose. When potential error can be quantified, it can often be accounted for in analysis. The bigger problem arises when error cannot be measured (or when we don’t even know it exists), because it cannot be corrected or factored into decision-making.

**Example:** If a GPS is accurate within ±5 m, that’s uncertainty you can plan for. If the GPS occasionally drifts 30 m without warning, that’s unknown error.

## Case Studies: Why Does Accuracy and Precision Matter

### Warming trends in the Southeast US

[Evidence](https://www.southernclimate.org/wp-content/uploads/NCA5_Ch22_Southeast.pdf) generally points to warming in the Southeast US. Researchers are using long-term temperature records from a network of weather stations to analyze trends. However, the stations vary in how often their sensors are calibrated, the precision of their measurements, and the completeness of their historical records. Consider the following questions

-   Which issues here relate to accuracy? Which relate to precision?

-   Imagine a research question where the dataset limits here **might not** be a problem. What would that question look like?

-   Imagine a research question where the same limits **might** cause a problem. What would that question look like?

### Landslide Risk

You are tasked with assessing landslide risk for a mountainous county: Your available data includes:

-   A digital elevation model
-   A soil type map
-   Rainfall records from a local station network

Consider the following questions:

-   Which parts of the dataset could lead to positional error, attribute error, or conceptual error?

-   Imagine a research question where the dataset limits here **might not** be a problem. What would that question look like?

-   Imagine a research question where the same dataset limits **might** cause a problem. What would that question look like?
